---
title: "04_Dimensionality_Reduction"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = FALSE
)
```

# Dimensionality Reduction

- As the name suggests, dimensionality reduction aims to reduce the number of separate dimensions in the data. This is possible because different genes are correlated if they are affected by the same biological process. Thus, we do not need to store separate information for individual genes, but can instead compress multiple features into a single dimension, e.g., an “eigengene”

```{r dataset_setup, message=FALSE}
#--- loading ---#
library(scRNAseq)
sce.zeisel <- ZeiselBrainData()

library(scater)
sce.zeisel <- aggregateAcrossFeatures(sce.zeisel, 
    id=sub("_loc[0-9]+$", "", rownames(sce.zeisel)))

#--- gene-annotation ---#
library(org.Mm.eg.db)
rowData(sce.zeisel)$Ensembl <- mapIds(org.Mm.eg.db, 
    keys=rownames(sce.zeisel), keytype="SYMBOL", column="ENSEMBL")

#--- quality-control ---#
stats <- perCellQCMetrics(sce.zeisel, subsets=list(
    Mt=rowData(sce.zeisel)$featureType=="mito"))
qc <- quickPerCellQC(stats, percent_subsets=c("altexps_ERCC_percent", 
    "subsets_Mt_percent"))
sce.zeisel <- sce.zeisel[,!qc$discard]

#--- normalization ---#
library(scran)
set.seed(1000)
clusters <- quickCluster(sce.zeisel)
sce.zeisel <- computeSumFactors(sce.zeisel, cluster=clusters) 
sce.zeisel <- logNormCounts(sce.zeisel)

#--- variance-modelling ---#
dec.zeisel <- modelGeneVarWithSpikes(sce.zeisel, "ERCC")
top.hvgs <- getTopHVGs(dec.zeisel, prop=0.1)
```

> mamba install -c conda-forge -c bioconda bioconductor-<package_name> # Installing a bioconductor package through mamba

```{r}
sce.zeisel
```

## Principal Component Analysis (PCA)

- the top principal components (PCs) capture the dominant factors of heterogeneity in the data set.

-  In the context of scRNA-seq, our assumption is that biological processes affect multiple genes in a coordinated manner.

- Mathematical Explanation: The use of the earlier PCs for denoising and data compaction is a strategy that is simple, highly effective and widely used in a variety of fields. It takes advantage of the well-studied theoretical properties of the PCA - namely, that a low-rank approximation formed from the top PCs is the optimal approximation of the original data for a given matrix rank. Indeed, the Euclidean distances between cells in PC space can be treated as an approximation of the same distances in the original dataset The literature for PCA also provides us with a range of fast implementations for scalable and efficient data analysis.

```{r}
library(scran)
top.zeisel <- getTopHVGs(dec.zeisel, n=2000)

set.seed(100) # See below.
sce.zeisel <- fixedPCA(sce.zeisel, subset.row=top.zeisel) 
reducedDimNames(sce.zeisel)
```

```{r}
dim(reducedDim(sce.zeisel, "PCA"))
```

- Using SVD approximation algorithms for faster and efficient results

```{r}
library(BiocSingular)
set.seed(1000)
sce.zeisel <- fixedPCA(sce.zeisel, subset.row=top.zeisel, 
    BSPARAM=RandomParam(), name="randomized")
reducedDimNames(sce.zeisel)
```

### Choosing number of Principal Components (PCs) to retain

- Using more PCs will retain more biological signal at the cost of including more noise that might mask said signal. On the other hand, using fewer PCs will introduce competition between different factors of variation, where weaker (but still interesting) factors may be pushed down into lower PCs and inadvertently discarded from downstream analyses.

- it is hard to determine whether an “optimal” choice exists for the number of PCs. Most practitioners will simply set d (no. of PCs) to a “reasonable” but arbitrary value, typically ranging from 10 to 50.

- The highest variation in the dataset is captured by the first few PCs

```{r variance_explained_by_PCs}
percent.var <- attr(reducedDim(sce.zeisel), "percentVar")
plot(percent.var, log="y", xlab="PC", ylab="Variance explained (%)")
```

### Visualizing the PCs

```{r}
library(scater)
plotReducedDim(sce.zeisel, dimred="PCA", colour_by="level1class")
```

- The problem is that PCA is a linear technique, i.e., only variation along a line in high-dimensional space is captured by each PC. As such, it cannot efficiently pack differences in  
d
  dimensions into the first 2 PCs.
  
- One workaround is to plot several of the top PCs against each other in pairwise plots (Figure 4.3). However, it is difficult to interpret multiple plots simultaneously, and even this approach is not sufficient to separate some of the annotated subpopulations.

```{r}
plotReducedDim(sce.zeisel, dimred="PCA", ncomponents=4,
    colour_by="level1class")
```


## Non-Linear methods for visualization

### t-Stochastic Neighbor Embedding

- The de facto standard for visualization of scRNA-seq data is the t -stochastic neighbor embedding (t-SNE) method (Van der Maaten and Hinton 2008).

- This attempts to find a low-dimensional representation of the data that preserves the distances between each point and its neighbors in the high-dimensional space.

```{r}
set.seed(00101001101)

# runTSNE() stores the t-SNE coordinates in the reducedDims
# for re-use across multiple plotReducedDim() calls.
sce.zeisel <- runTSNE(sce.zeisel, dimred="PCA")
plotReducedDim(sce.zeisel, dimred="TSNE", colour_by="level1class")
```

- Parameters to keep in mind for t-SNE: Random Initialization and Perplexity


```{r fig.height=6, fig.width=20}
set.seed(100)
sce.zeisel <- runTSNE(sce.zeisel, dimred="PCA", perplexity=5)
out5 <- plotReducedDim(sce.zeisel, dimred="TSNE",
    colour_by="level1class") + ggtitle("perplexity = 5")

set.seed(100)
sce.zeisel <- runTSNE(sce.zeisel, dimred="PCA", perplexity=20)
out20 <- plotReducedDim(sce.zeisel, dimred="TSNE",
    colour_by="level1class") + ggtitle("perplexity = 20")

set.seed(100)
sce.zeisel <- runTSNE(sce.zeisel, dimred="PCA", perplexity=80)
out80 <- plotReducedDim(sce.zeisel, dimred="TSNE", 
    colour_by="level1class") + ggtitle("perplexity = 80")

gridExtra::grid.arrange(out5, out20, out80, ncol=3)
```


- it is unwise to read too much into the relative sizes and positions of the visual clusters
- t-SNE is not obliged to preserve the relative locations of non-neighboring clusters, such that we cannot use their positions to determine relationships between distant clusters.
- t-SNE is proven tool for general-purpose visualization of scRNA-seq data and remains a popular choice in many analysis pipelines.

### Uniform Manifold Approximation and Projection (UMAP)

```{r}
set.seed(1100101001)
sce.zeisel <- runUMAP(sce.zeisel, dimred="PCA")
plotReducedDim(sce.zeisel, dimred="UMAP", colour_by="level1class")
```

- Compared to t-SNE, the UMAP visualization tends to have more compact visual clusters with more empty space between them. It also attempts to preserve more of the global structure than  t-SNE
- UMAP also involves a series of randomization steps so setting the seed is critical.
- the number of neighbors (n_neighbors) and the minimum distance between embedded points (min_dist) have the greatest effect on the granularity of the output.
- UMAP is unarguably much faster, and for that reason alone, it is increasingly displacing t-SNE as the method of choice for visualizing large scRNA-seq data sets.
- As a general rule, focusing on local neighborhoods provides the safest interpretation of t-SNE and UMAP plots.
 

















