---
title: "02_Getting_Datasets"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,   # This ensures the code is printed
  eval = FALSE   # This ensures no code chunks are actually executed
)
```


## Overview and Experimental Design

- In order to do statistical analysis on the data generated from a single cell experiment we need to first get the data into a usable format of **gene expression matrix** where the rows corresponds to gene IDs (ensembl) and the columns refers to individual cells (samples). Depending on the protocol the matrix might hold the counts of transcripts (SMART Seq 2) or counts of Unique Molecular Identifier (UMI) (10X Chromium). It is very important to have knowledge about the protocol that was used to generate a single cell RNA sequencing data set (either droplet based UMI or plate based UMI or reads) as the pre-processing of the raw data will depend on the protocol used.

- Different scRNA-seq protocols have their own advantages and disadvantages, however in today's research **droplet based technologies** are the defacto standard due to their low cost per cell and high throughput. While designing a scRNA-seq experiment how many cells to sequence and the sequencing depth varies on the biological question being asked and the design considerations for these experiments remain the same regarding batch effects and biological replicates. (power analysis can help?)

## Generating Counts Matrix From Raw Data

- In order to create pre-processing pipelines to analyze raw scRNA-seq data we first need to know the protocol that was used to generate that data as the upstream analysis (from RAW reads to Counts Matrix) is completely dependent on the sequencing protocol used, in contrast the downstream analysis (differential gene expression analysis, functional enrichment analysis, etc) is agnostic of the sequencing protocol. For example for data generated using droplet based technologies such as 10X Chromium we need to either use Cellranger software provided by 10X Genomics or a different pipeline built on bioconductor packages, where as for plate based read protocols such as SMART-Seq2 we can use similar pre-processing pipeline that is used for bulk-RNA Sequencing.


## Loading Processed Data into R

- The Bioconductor ecosystem uses a common data structure called **SingleCellExperiment** object for handling scRNA-seq data and facilitate interoperability across different packages within the ecosystem. The SingleCellExperiment object is made up of multiple slots as listed below,
a) rowData : To hold feature metadata (such as gene names)
b) assays : To hold the Primary and Transformed data (the gene expression matrix)
c) colData : To hold cell metadata (cell types etc)
b) reducedDims : To hold different reduced dimensions (such as PCA, UMAP and t-SNE)


- We can create SingleCellExperiment object from different data sources and formats. Some of the most common ones include: a) tabular files b) cellranger outputs (for 10X Chromium generated data) c) Hierarchical Data Format version 5 (HDF5) based formats such as H5AD. It is worth nothing that irrespective of the data source or format the analysis done on scRNA-seq data in R is almost always in standardized data structures such as SingleCellExperiment object or Seurat object from the Seurat ecosystem.

## Installing required packages

Since in my own setup (Ubuntu 25.10 dual boot) I am using mamba for package management and reproducibility installing packages in R could be a little tricky if it is done from the R console and hence it is recommended that installation of packages is done through mamba with the command shown below

```{bash, eval=FALSE}
# I am using a zsh shell so the shebang has zsh, if you are using bash then just replace the zsh with bash
!#bin/zsh

mamba install -c conda-forge r-biocmanager # Installs the repository manager for Bioconductor

mamba install -c conda-forge -c bioconda bioconductor-BiocFileCache # Rule of Thumb: Always use both channels in that specific order. bioconda relies on conda-forge for its foundation. Without conda-forge, the bio-packages are like software without an operating system.

```

> Install the required packages using mamba

```{bash}
mamba install -c conda-forge -c bioconda bioconductor-scuttle bioconductor-scran bioconductor-scater bioconductor-uwot bioconductor-rtracklayer bioconductor-BiocFileCache
```


## Loading data into R

```{r, message=FALSE, warning=FALSE}
# Load the required libraries
library(scuttle)
library(scran)
library(scater)
library(uwot)
library(rtracklayer)
library(BiocFileCache)


# Download file using BiocFileCache

bfc <- BiocFileCache("raw_data", ask = FALSE)
calero.counts <- bfcrpath(bfc, file.path("https://www.ebi.ac.uk/biostudies", 
    "files/E-MTAB-5522/counts_Calero_20160113.tsv"))
```

> Read in the matrix file and clean it

```{r}
mat <- read.delim(calero.counts, header=TRUE, row.names=1, check.names=FALSE)

# Only considering endogenous genes for now.
spike.mat <- mat[grepl("^ERCC-", rownames(mat)),] 
mat <- mat[grepl("^ENSMUSG", rownames(mat)),] 

# Splitting off the gene length column.
gene.length <- mat[,1]
mat <- as.matrix(mat[,-1]) 

dim(mat)
```


> Create SingleCellExperiment (sce) object using the matrix from previous step

```{r}
sce <- SingleCellExperiment(assays = list(counts = mat)) # Note that we provide our data as a named list where each entry of the list is a matrix - in this case, named "counts".

# What it basically means fram a data structure point of view is that the assay slot in an SCE object is a named list where each element corresponds to a matrix, for eg- the counts matrix is stored as an element in that list with the name as "counts" and the stored matrices can be accessed from the SCE object as shown below

# mat2 <- counts(sce) 
mat2 <- assay(sce, "counts") # most general method the first argument is the name of the SCE object while the second argument is the name of the element

```


> NOTE: For exported S4 classes, it is rarely a good idea to directly access the slots with the @ operator. This is considered bad practice as the class developers are free to alter the internal structure of the class, at which point any code using @ may no longer work. Rather, it is best to use the provided getter functions like assay() and counts() to extract data from the object. Similarly, setting slots should be done via the provided setter functions, which are discussed in the next section.


> Adding more representation of the primary data to the assay slot

```{r}
sce <- scuttle::logNormCounts(sce) # Computes the log normalized counts of the raw counts and stores it as another element in the named list

sce
```


**Example Output**
---
class: SingleCellExperiment 
dim: 46603 96 
metadata(0):
assays(2): counts logcounts
rownames(46603): ENSMUSG00000102693 ENSMUSG00000064842 ... ENSMUSG00000096730 ENSMUSG00000095742
rowData names(0):
colnames(96): SLX-9555.N701_S502.C89V9ANXX.s_1.r_1 SLX-9555.N701_S503.C89V9ANXX.s_1.r_1 ... SLX-9555.N712_S508.C89V9ANXX.s_1.r_1
  SLX-9555.N712_S517.C89V9ANXX.s_1.r_1
colData names(1): sizeFactor
reducedDimNames(0):
mainExpName: NULL
altExpNames(0):
---


```{r}
# Accessing the logcounts assay
norm_mat <- assay(sce, "logcounts")

# While the lognormcounts() function automatically returns the modified SCE object with the new assay (logcounts), when we want to do some custom transformation with the raw data we can also store that resultant matrix from the custom transformation in the assay slot as shown below

counts_100 <- counts(sce) + 100
assay(sce, "counts_100") <- counts_100 # assign a new entry to assays slot
assays(sce) # returns the total number of assays stored in this slot and the names of those

```

> **Important to remeber the getter() and setter() functions of different data structures to correctly modify the data objects**

```{r}
# Only keeping the first two assays
assays(sce) <- assays(sce)[1:2]

assayNames(sce) # returns the name of assays stored in the SCE object
names(assays(sce)) # same result, but slightly less efficient.
```


## Handling Metadata

To draw meaningful interpretation of the counts data we need to also handle the metadata correctly to understand and communicate the analysis effectively. Metadata such as sample names, cell type, batch number etc are all stored in the **colData** slot of an SCE object.

```{r}
# Downloading the SDRF file containing the metadata.
lun.sdrf <- bfcrpath(bfc, file.path("https://www.ebi.ac.uk/arrayexpress/files",
    "E-MTAB-5522/E-MTAB-5522.sdrf.txt"))
```

### Colum Metadata (Sample Metadata)

```{r}
coldata <- read.delim(lun.sdrf, check.names=FALSE)

# Only keeping the cells involved in the count matrix in 'mat'.
coldata <- coldata[coldata[,"Derived Array Data File"]=="counts_Calero_20160113.tsv",]

# Only keeping interesting columns, and setting the library names as the row names.
coldata <- DataFrame(
    genotype=coldata[,"Characteristics[genotype]"],
    phenotype=coldata[,"Characteristics[phenotype]"],
    spike_in=coldata[,"Factor Value[spike-in addition]"],
    row.names=coldata[,"Source Name"]
)

coldata
```


> We can either add the newly accquired colData to an existing SCE object or we can create a new SCE object from scratch as shown below,

```{r, eval=FALSE}
sce <- SingleCellExperiment(assays = list(counts=mat), colData=coldata) # from scratch
colData(sce) # view the stored colData

# adding colData to an existing SCE object
sce <- SingleCellExperiment(list(counts=mat))
colData(sce) <- coldata
```


**Regardless of which strategy you pick, it is very important to make sure that the rows of your colData actually refer to the same cells as the columns of your count matrix. It is usually a good idea to check that the names are consistent before constructing the SingleCellExperiment:**

```{r}
stopifnot(identical(rownames(coldata), colnames(mat)))
```


Some functions automatically add column metadata by returning a SingleCellExperiment with extra fields in the colData slot. For example, the scuttle package contains the addPerCellQC() function that appends a number of quality control metrics to the colData:

```{r}
sce <- scuttle::addPerCellQC(sce)
colData(sce)
```

### Row Metadata (Feature Metadata)

```{r}
rowData(sce)$Length <- gene.length
rowData(sce)
```

Some functions will return a SingleCellExperiment with the rowData populated with relevant bits of information, such as the addPerFeatureQC() function: **What does the per Feature QC metrics mean?**

```{r}
sce <- scuttle::addPerFeatureQC(sce)
rowData(sce)
```


> SCE object also has **rowRanges** slot to hold genomic co-ordinates *(why?)*

```{r}
# The paper uses Ensembl 82.
mm10.gtf <- bfcrpath(bfc, file.path("http://ftp.ensembl.org/pub/release-82",
    "gtf/mus_musculus/Mus_musculus.GRCm38.82.gtf.gz"))

gene.data <- rtracklayer::import(mm10.gtf)

# Cleaning up the object.
gene.data <- gene.data[gene.data$type=="gene"]
names(gene.data) <- gene.data$gene_id
is.gene.related <- grep("gene_", colnames(mcols(gene.data)))
mcols(gene.data) <- mcols(gene.data)[,is.gene.related]

rowRanges(sce) <- gene.data[rownames(sce)]
rowRanges(sce)[1:10,]
```































